% !TEX root = ../main.tex
%
% This is probably the only page that colleagues read, it is not necessary for the version sent to the reviewers, but it is usually written between the submission of the manuscript and the beginning of the slides, with a of emotion.
\pdfbookmark[0]{Remerciements}{Remerciements}
\addchap*{Remerciements}

A PhD is a long journey. For me, it took 3 years, 3 supervisors and education in 3 countries to finally reach here. Throughout this time, I have been supported by many people, without whom I would not have been able to reach here. I would like to take this opportunity to thank them. 

First and foremost, I would like to thank the rapporteurs of this thesis, Dr. David Bromberg (Professeur, Université de Rennes) and Dr. Pascal Felber (Professeur, Université de Neuchâtel). I am grateful for their time and effort in reviewing my work and providing valuable feedback. I would also like to thank the members of the jury, Dr. Gaël Thomas (Directeur de Recherche, INRIA), Dr. Sathya Peri (Professeur, IIT Hyderabad) and Dr. Stefania Dumbrava (Maîtresse de conférences, ENSIIE) for taking the time to read, evaluate and provide feedback on this body of work. The members of my CSD committee, Dr. Bernd Amann (Professeur, Sorbonne Université) and Dr. Ahmed Bouajjani (Professor, Université Paris Cité) have been a great source of support and guidance throughout my PhD. I am grateful for their time and effort in helping me navigate the challenges of a PhD and correcting the direction of my research. 

The people without whom this thesis would not have been possible are my supervisors, Marc Shapiro, Julien Sopena and Swan Dubois. From asking me if I would like to do a PhD back in 2021 while I was in Germany, to the day I defended my thesis, I cannot thank Marc enough for his constant support, guidance and well-placed reality checks. Julien, you have been a great inspiration, not only in the rigour of research but also in correcting my French which, even after several corrections, is still awful. Swan, you have been a great mentor and helped me navigate the finer details of theoretical computer science. I am grateful for the opportunity to have worked with you all.

To the seniors in the lab with whom I have shared numerous lunches, coffee breaks and discussions, thank you for making my time in the lab enjoyable. Thank you, Pierre, Jonathan, and Alessio for taking care of DELYS and your help in my research. To the members of the DELYS team, Dr. Luciana Arantes, Dr. Philippe Darche, Dr. Frank Petit; Thank you for keeping this amazing research group running.

I have had the pleasure of working with some amazing colleagues during my time in the lab. Saalik, Benoît and Laurent have been great friends and colleagues, helping me navigate french bureaucracy, teaching me bits of french culture and also helping me with my research direction. I am grateful for your friendship and support. To my fellow PhDs, Aymeric, Baptiste, Célia, Daniel and Étienne; Thank you for your discussions and support. I wish you well for your future. 

To the lehrstuhl softwaretechnik of RPTU, I extend my thanks, for introducing me to the wonderful world of parallel and distributed systems. I am grateful for the opportunity to have worked with Annette bieniusa, without whom I would not have been able to start my PhD.

Finally, the people without whom this thesis would not have been imaginable. My family, who dealt with years of daily phone and video calls, the fear of the COVID-19 pandemic, the uncertainty of being abroad and the stresses of a PhD, I am eternally grateful for your support. I dedicate this thesis to you Papa, Mummy and Yash. To my friends from France, Germany and India, thank you for your support, your love, and your friendship.

Thank you everyone. 